---
title: "River Chlorophyll-a Forecast Example"
output: html_document
---

# River Chlorophyll-a Forecast Example

This document is an example of how to submit a river chlorophyll-a forecast to the EFI-USGS river chlorophyll forecasting challenge using R.

## Source Necessary Packages

First, load the necessary packages for generating and submitting the forecast:

```{r}
library(tidyverse)
library(lubridate)
library(glue)

# The usgsrc4cast project has a few custom functions that are needed to successfully submit to the EFI-USGS river chlorophyll forecast challenge 
source("https://raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/prod/R/eco4cast-helpers/submit.R")
source("https://raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/prod/R/eco4cast-helpers/forecast_output_validator.R")
source("https://raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/prod/R/eco4cast-helpers/noaa_gefs.R")
```

## Step 0: Model Configuration
Define a unique name to identify your model in the leaderboard and connect it to team members' info. If running an example model, include the word "example" in your model_id - any model with "example" in the name will not be scored in the challenge.
```{r}
model_id <- "usgsrc4cast_example"
project_id <- "usgsrc4cast"
forecast_duration <- "P1D"
target_var <- "chla"
```

## Step 1: Download Latest Target Data and Site Description Data
Focus on the focal sites for this example. When you're comfortable with your model and workflow, you can expand to all sites in the challenge.

```{r}
focal_sites <- c("USGS-05553700", "USGS-14211720")

site_data <- readr::read_csv("https://raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/prod/USGS_site_metadata.csv") %>% 
  filter(site_id %in% focal_sites)

target <- readr::read_csv("https://sdsc.osn.xsede.org/bio230014-bucket01/challenges/targets/project_id=usgsrc4cast/duration=P1D/river-chl-targets.csv.gz") %>%
  dplyr::filter(site_id %in% focal_sites) %>% 
  dplyr::distinct()
```

## Step 2: Get Meteorological Predictions as Drivers
Get the latest forecast date and NOAA weather forecasts.
```{r}
forecast_date <- Sys.Date()
noaa_date <- Sys.Date() - days(2)

noaa_historic <- noaa_stage3(project_id = project_id)

noaa_mean_historic <- noaa_historic %>% 
    dplyr::filter(site_id %in% focal_sites) %>% 
    dplyr::rename(ensemble = parameter) %>% 
    dplyr::select(site_id, datetime, prediction, ensemble, variable) %>% 
    dplyr::mutate(date = as_date(datetime)) %>% 
    dplyr::group_by(site_id, date, variable) %>% 
    dplyr::summarize(predicted_mean = mean(prediction, na.rm = TRUE),
                     .groups = "drop") %>% 
    dplyr::rename(datetime = date) %>% 
    dplyr::collect()

noaa_forecast <- noaa_stage2(project_id = project_id, start_date = noaa_date)

noaa_mean_forecast <- noaa_forecast %>% 
  dplyr::filter(site_id %in% focal_sites, reference_datetime == lubridate::as_datetime(noaa_date)) %>% 
  dplyr::rename(ensemble = parameter) %>% 
  dplyr::select(site_id, reference_datetime, datetime, variable, prediction, ensemble) %>% 
  dplyr::mutate(date = as_date(datetime)) %>% 
  dplyr::group_by(site_id, reference_datetime, date, variable, ensemble) %>% 
  dplyr::summarize(predicted_mean = mean(prediction, na.rm = TRUE),
                   .groups = "drop") %>% 
  dplyr::rename(datetime = date) %>% 
  dplyr::collect()
```

## Step 3: Define the Forecasts Model for a Site
Run the model for each site and generate forecasts.
```{r}
all_forecasts <- c()

for(site in focal_sites){
  message(paste0("Running site: ", site))
  
  cur_noaa_historic <- filter(noaa_mean_historic, site_id == site) %>% 
    pivot_wider(names_from = variable, values_from = predicted_mean)
  
  site_target <- target %>% 
    dplyr::select(datetime, site_id, variable, observation) %>% 
    dplyr::filter(variable == "chla", site_id == site) %>% 
    tidyr::pivot_wider(names_from = "variable", values_from = "observation") %>% 
    dplyr::left_join(cur_noaa_historic, by = c("datetime", "site_id")) %>% 
    filter_at(3:ncol(.), all_vars(!is.na(.))) %>% 
    mutate(chla_lagged_1 = dplyr::lag(chla, n = 1)) %>%
    slice(2:n())
  
  fit <- lm(chla ~ chla_lagged_1 + air_temperature, data = site_target)
  
  lagged_chla <- tibble(datetime = forecast_date, 
                        chla_lagged_1 = ifelse((forecast_date - 1) %in% site_target$datetime,
                                               site_target$chla[site_target$datetime == (forecast_date - 1)],
                                               NA)) %>% 
    mutate(chla_lagged_1 = ifelse(is.na(chla_lagged_1),
                                  mean(site_target$chla, na.rm = TRUE),
                                  chla_lagged_1))
    
  cur_noaa_forecast <- filter(noaa_mean_forecast, site_id == site, datetime >= forecast_date, variable == 'air_temperature') %>% 
    select(-reference_datetime) %>% 
    pivot_wider(names_from = variable, values_from = predicted_mean) %>% 
    left_join(select(lagged_chla, datetime, chla_lagged_1))
  
  forecasted_dates <- sort(lubridate::date(unique(cur_noaa_forecast$datetime)))
  ensembles <- sort(unique(cur_noaa_forecast$ensemble))
  
  forecasted_chla <- c()
  
  for(i in 1:length(forecasted_dates)){
    valid_date <- forecasted_dates[i]
    if(i == 1){
      cur_data <- filter(cur_noaa_forecast, datetime == valid_date) %>% 
        mutate(chla = predict(fit, tibble(air_temperature, chla_lagged_1)))
    } else {
      cur_lagged <- filter(forecasted_chla, datetime == (valid_date - 1)) %>% 
        mutate(datetime = valid_date) %>% 
        rename(chla_lagged_1 = chla)
      cur_data <- filter(cur_noaa_forecast, datetime == valid_date) %>% 
        select(-chla_lagged_1) %>%  
        left_join(cur_lagged, by = c("site_id", "datetime", "ensemble")) %>% 
        mutate(chla = predict(fit, tibble(air_temperature, chla_lagged_1)))
    }

    forecasted_chla <- bind_rows(forecasted_chla, 
                                 select(cur_data, site_id, datetime, ensemble, chla))
  }
  
  forecasted_chla <- forecasted_chla %>% 
    mutate(reference_datetime = forecast_date,
           family = "ensemble",
           variable = "chla", 
           duration = "P1D", 
           project_id = project_id, 
           model_id = model_id) %>% 
    rename(parameter = ensemble,
           prediction = chla) %>% 
    select(project_id, model_id, datetime, reference_datetime,
           site_id, duration, family, parameter, variable, prediction)
  
  all_forecasts <- bind_rows(all_forecasts, forecasted_chla) 
}
```

### Visualize the Predictions
```{r}
all_forecasts %>% 
  ggplot(aes(x = datetime, y = prediction, group = parameter)) +
  geom_line(alpha=0.3) +
  facet_wrap(~site_id, scales = "free")

```

### Forecast Output File
Format the forecast output file name and write the CSV to disk.
```{r}
file_date <- forecast_date
forecast_file <- paste0("usgsrc4cast", "-", file_date, "-", model_id, ".csv")

write_csv(all_forecasts, forecast_file)

```

## Step 4: Submit Forecast
Submit the forecast.
```{r}
submit(forecast_file = forecast_file, 
       project_id = project_id, 
       ask = FALSE)

```

With this example, you should be able to generate and submit a river chlorophyll-a forecast to the challenge. Adjust the model_id and use your own model to generate forecasts.

